% \documentclass{article}

% \begin{document}

\section{The task in greater detail}
\label{sec:task}


\subsection{The domain}
\label{sec:domain}
Hofstadter's interest in sequence-extending programs goes back at least a quarter of a centuary, and the tasks intended for ``the program'' have shifted over time. In the original pre-Seek-Whence guise, a typical input for the program could have been, for instance, the sequence of alternate primes \emph{2 5 11 17 23$\ldots$}.

Gradually Hofstadter's interest shifted away from such mathematical-knowledge intensive sequences to a smaller domain: that of sequences which involve no knowledge of arithmetics beyond the relationship of successor and predecessor, and the ability to count. This meant outlawing such sequences as the primes, and even the addition-involving Fibonacci sequence. The shift to this stark domain is described in \citeN[pages 48-49]{Hofstadter:FCCA}.

This restricted domain is by no means poor in sequences that tax the intellect, or even in sequences excellent for exploring program-cognition. For sequence \emph{k} in the next section, for example, it is interesting to contemplate how a program might navigate the \emph{2 2 2 2 2} conundrum. Several other sequences require for their honest human-like understanding the ability to chunk (almost all examples in the next section) and even to see an object \emph{as} another (examples \emph{e, f, g, } and \emph{h}).

I shall not here try to delineate further the problems that could be the inputs for Seqsee, but leave that task instead to the dozen or so examples in the next section.

\subsection{A wish-list of sequences}
\label{sec:list}

The following annotated listing of sequences is offered in ordered to add accountability: These are sequences that Seqsee should ``get'' before we can pronounce it to be \emph{good enough}. The list starts out rather simply, but builds tempo as it goes along.

Most of these sequences come from chapter 1 of \citeN{Hofstadter:FCCA}.

\renewcommand{\theenumi}{\alph{enumi}}
\begin{enumerate}
\item \textbf{1 2 3 4 $\ldots$}

Seqsee must get this, of course, but it should also get this rapidly even if a hundred terms are initially presented. It's performance should be no worse than linear. If it takes time quadratic (or even just super-linear) in the number of input terms, there is something seriously wrong with the model.

\item \textbf{ 1 1 2 2 3 3 $\ldots$}
\item \textbf{ 1 2 2 3 3 3 $\ldots$}

Seqsee should see the initial ``group of size 1''.

\item \textbf{12 13 14 11 12 13 14 10 11 12 13 14 $\ldots$}

The program---and it is aware of negative numbers---should know that it is going to hit a wall in the future.

\item \textbf{1 1 2 3 1 2 2 3 4 1 2 3 3 4 5 $\ldots$}

I want Seqsee to be \emph{able} to see the initial 1123 as a variant of 123, and likewise the following groups as variants of 1234 and 12345.

\item \textbf{1 1 2 3 1 2 2 3 1 2 3 3 1 1 2 3 1 2 2 3 1 2 3 3$\ldots$}
\item \textbf{1 1 2 3 1 2 2 3 1 2 3 3 1 2 2 3 1 1 2 3 1 2 2 3$\ldots$}
\item \textbf{1 1 2 3 1 2 2 3 1 2 3 3 1 2 3 1 2 3 1 2 3$\ldots$}

These three sequences are all ``doubler'' sequences: the first two were called marching- and bouncing-doubler, respectively, by Hofstadter. I call the last the piercing-doubler.

As in the sequence \emph{e} above, Seqsee should be able to see an endless list of 123 variants, with funny things happening in all three cases. As the names of these sequences indicate, people can perceive ``motion'' happening in these, and I'd like the program to be able to, too. It should also notice that all three are similar sequences, cut from the same cloth.

\item \textbf{ 2 10 1 3 11 2 4 12 3$\ldots$}
\item \textbf{ 2 10 2 3 11 2 3 4 12$\ldots$}

These interlaced sequences cannot be perceived merely by a rudimentary ``try every n$^\mathrm{th}$ term and see if they fit together'', as can be seen from sequence \emph{j}.

\item \textbf{ 2 1 2 2 2 2 2 3 2 2 4 2$\ldots$}

A confounding cousin of \emph{i} and \emph{j} above.

\item \textbf{ 1 0 0 1 1 1 0 0 0 0$\ldots$}
\item \textbf{ 1 2 3 1 2 2 3 1 2 2 2 3$\ldots$}
\item \textbf{ 5 4 3 4 5 4 3 4 5 $\ldots$}
\item \textbf{ 1 2 3 3 4 4 5 5 5 6 6 6$\ldots$}

Somehow I feel that the last one above is going to require the greatest amount of work.
\end{enumerate}

\subsection{More than just extending}
\label{sec:more}
The goals of the project also include things apart from extending.  Making variations on a theme is an example of such a goal.  To take a simple example a variation of \emph{1 2 3 1 2 3 1 2 3} is \emph{1 2 3 4 1 2 3 4 1 2 3 4}.  A more subtle example would be the variation of  sequence f (the marching doubler) to produce the marching singler (\emph{1 2 2 3 3 1 1 2 3 3 1 1 2 2 3}). \citeN[chapter 12]{Hofstadter:Meta} discusses the general theme of variation making.

It is important to note that how we perceive the sequence influences the variations we make.  The marching doubler can be perceived in a life-less way as the following 12 terms repeating indefinitely: \emph{1 1 2 3 1 2 2 3 1 2 3 3}.  seen this way without noticing the internal structure a faithful variant would be the following 12 terms repeated indefinitely: \emph{17 2 11 18 5 1 6 9 7 12 5 13}. In some sense that would completely miss the point. Thus the issue of variation making is not divorced from the issue of sequence perception.

Another aim is that the program be reminded of earlier sequences that were similar.  This functionality is similar to the episodic memory in Metacat \cite{Marshall}.

A third aspect of going beyond just extending concerns meta-thoughts.  Consider the question of when to stop.  Whether we have found the answer or not is a judgment call, as is the question of whether we should give up. A central question that Metacat addressed beyond Copycat was that of self-watching, and self-watching will also be very important for Seqsee. It already has some meta-thoughts. Now that the other things are falling into place I can pay self-watching and meta-thoughts the attention that is due. 

Most tasks in this section belong to the category \emph{still to be done}, but they are absolutely crucial to the project, and would get a large chunk of my efforts in the coming days.
